---
title: "Predicting the Next Big Earthquake"
subtitle: "A Bayesian Analysis of M8+ Inter-Event Times"
author: "Group 8: Rean Du, Oliver Chu, Daniel Ernesto Santos, Keerthana Girish"
date: "STATS 115 — Winter 2026"
format:
  revealjs:
    theme: default
    slide-number: true
    smaller: true
---

```{r setup, include=FALSE}
library(dplyr)
library(rstan)
library(tidyverse)
library(bayesrules)
library(MASS)
library(rstanarm)
library(ggplot2)
library(bayesplot)
```


```{r loadData, include=FALSE}
prior_eqs <- read_csv("data/earthquakes_before_1950.csv")

prior <- prior_eqs[, c("time", "mag")]
prior$datetime <- as.POSIXct(prior$time, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC")
prior <- prior[order(prior$datetime), ]
prior$intertime <- c(NA, as.numeric(
  difftime(prior$datetime[-1],prior$datetime[-nrow(prior)], units="days")))
prior$Display <- round(prior$intertime, 2)

between_time <- prior$intertime[!is.na(prior$intertime)]
E_T <- mean(between_time)
shape <- 2
rate <- 2 / E_T

theta_exp <- rate
theta_var <- theta_exp^2 / 5
b <- theta_exp / theta_var
a <- theta_exp * b

data <- read.csv("data/earthquake.csv")
earthquake <- data[, c("time", "mag")]
earthquake$datetime <- as.POSIXct(earthquake$time, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC")
earthquake <- earthquake[order(earthquake$datetime), ]
earthquake$intertime <- c(NA, as.numeric(
  difftime(earthquake$datetime[-1],earthquake$datetime[-nrow(earthquake)], units="days")))
earthquake$Display <- round(earthquake$intertime, 2)

tObs <- earthquake$intertime[!is.na(earthquake$intertime)]
n <- length(tObs)
sum_t <- sum(earthquake$intertime, na.rm = TRUE)
lastDate <- max(earthquake$datetime)
xDay <- as.numeric(difftime(Sys.Date(), lastDate, units="days"))

aPost <- a + 2*n
bPost <- b + sum_t

model <- "
data {
  int<lower=1> N;
  vector<lower=0>[N] T;
}
parameters {
  real<lower=0> theta;
}
model {
  T ~ gamma(2, theta);
  theta ~ gamma(5.0, 1120.523);
}
"

simulation <- stan(model_code=model, data=list(N = n, T = tObs),
                   chains=4, iter=5000*2, seed=1234,
                   refresh=FALSE)

postTheta <- as.data.frame(simulation, pars = "theta")$theta
priorTheta <- rgamma(10000, shape = a, rate = b)
cred_int_95 <- quantile(postTheta, probs = c(0.025, 0.975))

remainDays <- (xDay + 2/postTheta) / (1 + postTheta * xDay)

check_sim <- as.data.frame(simulation)
check_sim <- check_sim %>%
  mutate(y_predict = rgamma(n = n(), shape = 2, rate = theta))
check_sim <- check_sim %>%
  mutate(days_from_today = y_predict - xDay)
check_sim_future <- check_sim %>%
  filter(days_from_today > 0)
```


## Motivation

Earthquakes of magnitude 8 or above occur roughly once per year worldwide. Despite their rarity, they account for a disproportionate share of seismic fatalities and infrastructure damage. Even coarse probabilistic forecasts of when the next event might occur can inform resource allocation, preparedness planning, and insurance risk modeling.

::: {.columns}
::: {.column width="55%"}
We address two questions using Bayesian inference on historical USGS data:

**Q1.** Given `r round(xDay)` days have elapsed since the last M8+ earthquake, what is the expected remaining waiting time?

**Q2.** What is the probability the next M8+ earthquake occurs within $d$ days from now?

Both require conditioning on elapsed time, because tectonic stress accumulates — the longer the wait, the higher the hazard.
:::
::: {.column width="45%"}
[PLACEHOLDER: World map or timeline graphic showing recent M8+ earthquakes]
:::
:::


## Data

::: {.columns}
::: {.column width="50%"}
We use the USGS Earthquake Catalog filtered to $M \geq 8.0$, converted to inter-event times (days between consecutive earthquakes).

**Pre-1950** (39 events, 38 inter-event times): used to construct the informative prior for $\theta$. Mean inter-event time: `r round(E_T, 1)` days.

**Post-1950** (`r n + 1` events, `r n` inter-event times): observed data for the likelihood. Most recent event: `r format(lastDate, "%B %d, %Y")`. Days elapsed since: `r round(xDay)`.
:::
::: {.column width="50%"}
```{r hist-intertime, echo=FALSE, fig.height=5, fig.width=5}
ggplot(data.frame(t = tObs), aes(x = t)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20,
                 fill = "steelblue", color = "white", alpha = 0.7) +
  labs(title = "Post-1950 Inter-Event Times",
       subtitle = paste0("n = ", n, " intervals between consecutive M8+ earthquakes"),
       x = "Days Between M8+ Earthquakes",
       y = "Density") +
  theme_minimal()
```
:::
:::


## Model: Gamma(2, θ)

We model each inter-event time as $T_i \mid \theta \sim \text{Gamma}(2, \theta)$, where $\alpha = 2$ represents a two-stage stress accumulation process and $\theta$ is the rate at which each stage completes. The Exponential ($\alpha = 1$) is memoryless: elapsed time would carry no information. The Gamma with $\alpha = 2$ has an increasing hazard rate, so the longer we wait, the more likely the next event becomes. This matches the physical intuition that tectonic stress builds over time.

The prior for $\theta$ is constructed from pre-1950 data. We set $\hat\theta = 2/E(T) \approx `r round(rate, 5)`$ and encode it as $\theta \sim \text{Gamma}(`r round(a, 1)`,\; `r round(b, 1)`)$, choosing a variance of $\hat\theta^2 / 5$ to keep the prior informative but not dominant over the observed data.

::: aside
^[PLACEHOLDER: Source]
:::


## Posterior Derivation

The Gamma prior is conjugate to the Gamma likelihood:

| | Expression |
|---|---|
| **Prior** | $\theta \sim \text{Gamma}(a,\; b)$ |
| **Likelihood** | $L(\theta) \propto \theta^{2n}\, e^{-\theta \sum t_i}$ |
| **Posterior** | $\theta \mid \mathbf{t} \sim \text{Gamma}(a + 2n,\; b + \sum t_i)$ |

Substituting: $\;\theta \mid \mathbf{t} \sim \text{Gamma}(`r round(aPost, 1)`,\; `r round(bPost, 1)`)$. We sample from this posterior using Stan MCMC with 4 chains and 10,000 iterations each.


## MCMC Diagnostics

::: {.columns}
::: {.column width="50%"}
```{r trace-plot, echo=FALSE, fig.height=4, fig.width=5}
bayesplot::mcmc_trace(simulation, pars = "theta", size = 0.5) +
  labs(title = "Trace Plot",
       subtitle = "Four chains sampling θ; good mixing indicates convergence")
```
:::
::: {.column width="50%"}
```{r density-overlay, echo=FALSE, fig.height=4, fig.width=5}
bayesplot::mcmc_dens_overlay(simulation, pars = "theta") +
  labs(title = "Density by Chain",
       subtitle = "Overlapping densities confirm agreement across chains")
```
:::
:::

All four chains mix well. $\hat{R} \approx 1.0003$ (threshold: $<1.01$), effective sample size $\approx 6{,}987$.


## Prior vs. Posterior

::: {.columns}
::: {.column width="50%"}
```{r prior-vs-post, echo=FALSE, fig.height=5, fig.width=5}
plot(density(priorTheta), lwd = 3, col = "black",
     xlab = expression(theta), ylab = "Density",
     main = expression("Prior (black) and Posterior (red) of " * theta),
     ylim = c(0, 1100), xlim = c(0, 0.016))
lines(density(postTheta), lwd = 3, col = "red")
legend("topright", legend = c("Prior", "Posterior"),
       col = c("black", "red"), lwd = 3, cex = 0.9)
mtext(paste0("The ", n, " observed inter-event times narrow the prior substantially."),
      side = 1, line = 4, cex = 0.7)
```
:::
::: {.column width="50%"}

| Statistic | Value |
|----------|-------|
| Posterior mean | `r round(mean(postTheta), 5)` |
| Posterior median | `r round(median(postTheta), 5)` |
| 95% credible interval | (`r round(cred_int_95[1], 5)`, `r round(cred_int_95[2], 5)`) |
| Implied mean recurrence | `r round(2 / mean(postTheta))` days ($\approx$ `r round(2 / mean(postTheta) / 365, 1)` yr) |

The `r n` observed inter-event times concentrate the posterior around $\theta \approx 0.0044$, substantially narrowing the prior. At this value, the expected time between consecutive M8+ earthquakes is roughly `r round(2 / mean(postTheta) / 365, 1)` years.
:::
:::


## Conditional Prediction: Expected Remaining Time

Because Gamma(2, $\theta$) is not memoryless, we condition on elapsed time $x$:

$$E[T - x \mid T > x,\; \theta] = \frac{x + \frac{2}{\theta}}{1 + \theta x}$$

::: {.columns}
::: {.column width="50%"}
```{r remain-density, echo=FALSE, fig.height=4.5, fig.width=5}
df_remain <- data.frame(days = remainDays)
ggplot(df_remain, aes(x = days)) +
  geom_density(fill = "steelblue", alpha = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = mean(remainDays), color = "red", linewidth = 1) +
  labs(title = "Posterior Predictive Remaining Wait",
       subtitle = paste0("Conditioned on x = ", round(xDay),
                         " days elapsed | Red line: mean = ",
                         round(mean(remainDays)), " days"),
       x = "Remaining Days", y = "Density") +
  theme_minimal()
```
:::
::: {.column width="50%"}

| Quantile | Remaining Days |
|----------|---------------|
| Min | `r round(min(remainDays))` |
| 25th | `r round(quantile(remainDays, 0.25))` |
| Median | `r round(median(remainDays))` |
| 75th | `r round(quantile(remainDays, 0.75))` |
| Max | `r round(max(remainDays))` |

With `r round(xDay)` days elapsed, the expected remaining wait is **~`r round(mean(remainDays))` days**, averaged over the full posterior of $\theta$.
:::
:::


## Conditional Prediction: P(Earthquake Within $d$ Days)

The probability that the next M8+ earthquake occurs within $d$ days from now, given $x$ days elapsed:

$$P(T \leq x + d \mid T > x,\; \theta) = \frac{F(x + d \mid \theta) - F(x \mid \theta)}{1 - F(x \mid \theta)}$$

We average over the posterior of $\theta$:

```{r prob-table, echo=FALSE}
daysCheck <- c(100, 200, 300, 365, 500, 730)

probs <- sapply(daysCheck, function(d) {
  probInDays <- (
    pgamma(xDay + d, shape = 2, rate = postTheta) -
    pgamma(xDay, shape = 2, rate = postTheta)
  ) / (
    1 - pgamma(xDay, shape = 2, rate = postTheta)
  )
  mean(probInDays)
})

prob_df <- data.frame(
  `Days from now (d)` = daysCheck,
  `Months` = round(daysCheck / 30.44, 1),
  `P(next quake within d days)` = paste0(round(probs * 100, 1), "%"),
  check.names = FALSE
)

knitr::kable(prob_df, align = "c",
             caption = paste0("Conditional probabilities given x = ", round(xDay), " days elapsed since last M8+ event."))
```

The increasing probabilities reflect the non-memoryless property: as elapsed time grows, the conditional hazard rises.


## Model Validation

::: {.columns}
::: {.column width="50%"}
```{r model-check, echo=FALSE, fig.height=5, fig.width=5}
ggplot() +
  geom_density(aes(x = tObs, color = "Observed"), linewidth = 1) +
  geom_density(aes(x = check_sim$y_predict, color = "Posterior Predictive"), linewidth = 1) +
  labs(title = "Observed vs. Posterior Predictive",
       subtitle = "Close agreement supports the Gamma(2, θ) specification",
       x = "Days Between Earthquakes", y = "Density") +
  scale_color_manual(values = c("Observed" = "black", "Posterior Predictive" = "red")) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")
```
:::
::: {.column width="50%"}
The posterior predictive distribution closely tracks the empirical distribution of the `r n` observed inter-event times. The slight rightward shift reflects posterior uncertainty — the model hedges toward longer wait times rather than being overconfident. This supports Gamma(2, $\theta$) as a reasonable specification for M8+ earthquake recurrence.
:::
:::


## Conclusion

Given `r round(xDay)` days since the last M8+ earthquake, the model estimates an expected remaining wait of approximately `r round(mean(remainDays))` days, with a 95% credible interval for $\theta$ of (`r round(cred_int_95[1], 4)`, `r round(cred_int_95[2], 4)`).

The conditional probability framework is central to this analysis. Because Gamma(2, $\theta$) has an increasing hazard rate, the probability of an earthquake within any fixed window grows as elapsed time increases. The posterior predictive closely matches observed data, supporting the model specification.

**Limitations.** The model assumes independence between events, a fixed shape parameter ($\alpha = 2$), and a stationary rate. These simplifications are standard for a first-order analysis but would need relaxation for operational forecasting.
