---
title: "STATS 115 Final Project: Predicting the Next Big Earthquake"
author: "Group 8: Rean Du, Oliver Chu, Daniel Ernesto Santos, Keerthana Girish" 
format: revealjs
---

```{r setup, include=FALSE}
library(dplyr)
library(rstan)
library(tidyverse)
library(bayesrules)
library(MASS)
library(rstanarm)
library(ggplot2)
library(bayesplot)
```


```{r loadData, include=FALSE}
prior_eqs <- read_csv("data/earthquakes_before_1950.csv")

prior <- prior_eqs[, c("time", "mag")]
prior$datetime <- as.POSIXct(prior$time, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC")
prior <- prior[order(prior$datetime), ]
prior$intertime <- c(NA, as.numeric(
  difftime(prior$datetime[-1],prior$datetime[-nrow(prior)], units="days")))
prior$Display <- round(prior$intertime, 2)
```


## Introduction to the Project

- Large earthquakes are rare but extremely destructive
- Predicting *when* the next one might occur is very difficult
- Even rough estimates can help with:
  - disaster preparedness
  - infrastructure planning
  - risk assessment

Is it possible to use past earthquake data to predict when the next one could happen?


## Our Project Goal (Proposal Question)

-  What is the expected waiting time until the next earthquake with magnitude above 8 happens worldwide given observed inter-event times?
-  What is the probability that the next earthquake will occur within a specific time window?


## Data

- Global earthquakes with magnitude Ōēź 8 taken from USGS website
- Converted into inter-event times (time between earthquakes)
- Split into:
  - Pre-1950 ŌåÆ used for prior
  - Post-1950 ŌåÆ used for likelihood

## Challenges
- Small number of observations (rare events)
- Potential missing or inconsistent historical records
- Assumes independence between events


## Model Choice
We decided to model inter-event times as $T_i$:
$$T_i|\theta \sim Gamma(2, \theta)$$
-  $T_i$: waiting time (in days) between 2 earthquakes  
-   2: number of stress accumulation stages  
-   $\theta$: scale that shows the duration of each stage  

### Why Gamma?
- Gamma can model waiting times
- Flexible shape
- Captures multi-stage processes

## Prior for ╬Ė

- Built using pre-1950 earthquake data
- Shape fixed at 2
- Estimated rate from mean inter-event time -> mean value of all inter-event times = E(T)
  - $\theta = \frac{2}{E(T)}$

- Prior reflects historical seismic behavior before modern data
- Used expected value $\theta$ to find the shape and rate of gamma distribution for $'theta$: $\theta \sim Gamma(a, b)$


## Bayesian Model

Prior:
$$
\theta \sim Gamma(a,b)
$$

Likelihood:
$$
T_i | \theta \sim Gamma(2,\theta)
$$

Posterior:
$$
p(\theta|T) \propto p(T|\theta)p(\theta)
$$

## Results of MCMC

## Predictions

## Interpretation

## Future possibilities

- Model earthquakes by region and fault line
- Use hierarchical Bayesian models
- Include time-varying behavior

Goal: build a more realistic and predictive model

## Conclusion

